---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# gghdx

<!-- badges: start -->
[![R-CMD-check](https://github.com/caldwellst/gghdx/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/caldwellst/gghdx/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/caldwellst/gghdx/branch/main/graph/badge.svg)](https://app.codecov.io/gh/caldwellst/gghdx?branch=main)
<!-- badges: end -->

The goal of gghdx is to make it as simple as possible to follow the HDX visual
guidelines when creating graphs using ggplot2. While most of the functionality
is in allowing easy application of the HDX color ramps, the package also
streamlines some of the other recommendations and best practices regarding
plotted text, axis gridlines, and other visual features.

## Installation

You can install gghdx directly from GitHub:

```r
## install.packages("remotes")
remotes::install_github("caldwellst/gghdx")
```

The package is not currently available on CRAN. Since gghdx utilizes the
showtext and sysfonts packages to cleanly import and use the Source Sans Pro
font, it is recommended to install them as well. Simply just install showtext
and sysfonts will also be installed.

```r
install.packages("showtext")
```

The package is designed so the user just has to run `gghdx()` once a session
and mainly forget about it. This will automatically set your ggplot2 defaults
to use the HDX theme, palettes, fonts, and more. If you want more control or
want to better understand how the package works, please see the details below!

```{r child="vignettes/gghdx.Rmd"}
```
